---
title: "18S basic alpha, beta diversity, and taxa composition"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
.libPaths()
loadedNamespaces()
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
library(phyloseq)
#devtools::install_github("jbisanz/qiime2R")
library(qiime2R)
library(vegan)
#devtools::install_github("gauravsk/ranacapa")
library(ranacapa)
library(patchwork)
library(scales)
library(psadd)
```

# Paths

```{r paths}
raw_dir <- "Diet_data"

feature_qza  <- file.path(raw_dir, "table.qza")
taxonomy_qza <- file.path(raw_dir, "taxonomy.qza")
metadata_txt <- file.path(raw_dir, "sample-metadata.txt")
tree_qza      <- file.path(raw_dir, "rooted-tree.qza")

fig_dir <- "Figures_Diet"
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
```

# Metadata

The metadata should be saved as `Raw data/sample-metadata_18S.txt` with columns:

```text
sampleid	site
Shikine_C1	Control
...
Shikine_H16	High-CO2
```

```{r metadata-import}
metadata <- readr::read_tsv(metadata_txt, comment = "", show_col_types = FALSE) %>%
  mutate(
    sampleid = as.character(sampleid),
    site = factor(site, levels = c("Control", "High-CO2"))
  )

metadata %>% count(site)
stopifnot(!any(is.na(metadata$site)))
```

# Import 18S data

```{r import}
if(file.exists(tree_qza)){
  ps <- qza_to_phyloseq(
    features = feature_qza,
    taxonomy = taxonomy_qza,
    tree     = tree_qza
  )
} else {
  ps <- qza_to_phyloseq(
    features = feature_qza,
    taxonomy = taxonomy_qza
  )
}

# Keep only samples present in the metadata and attach metadata manually.
metadata_keep <- metadata %>% filter(sampleid %in% sample_names(ps))
missing_in_phyloseq <- setdiff(metadata$sampleid, sample_names(ps))
missing_in_metadata <- setdiff(sample_names(ps), metadata$sampleid)

if(length(missing_in_phyloseq) > 0){
  message("Samples in metadata but not in feature table: ", paste(missing_in_phyloseq, collapse = ", "))
}
if(length(missing_in_metadata) > 0){
  message("Samples in feature table but not in metadata: ", paste(missing_in_metadata, collapse = ", "))
}
```

```{r}

host_taxa_patterns <- c("Echinoidea")
host_pattern <- paste(host_taxa_patterns, collapse = "|")

tax_df <- as.data.frame(tax_table(ps), stringsAsFactors = FALSE) %>%
  mutate(across(everything(), as.character))

host_taxa <- rownames(tax_df)[
  apply(tax_df, 1, function(x){
    any(stringr::str_detect(
      stringr::str_to_lower(x),
      stringr::str_to_lower(host_pattern)
    ), na.rm = TRUE)
  })
]

host_filter_summary <- tibble(
  n_asvs_removed = length(host_taxa),
  reads_removed = sum(taxa_sums(ps)[host_taxa]),
  reads_before_filter = sum(sample_sums(ps)),
  percent_reads_removed = 100 * reads_removed / reads_before_filter
)

host_taxa_table <- tax_df[host_taxa, , drop = FALSE] %>%
  rownames_to_column("ASV")

write_csv(host_filter_summary, file.path(fig_dir, "host_echinoidea_filter_summary_18S.csv"))
write_csv(host_taxa_table, file.path(fig_dir, "host_echinoidea_removed_taxa_18S.csv"))

ps <- prune_taxa(!taxa_names(ps) %in% host_taxa, ps)
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)

```


```{r import}
ps <- prune_samples(metadata_keep$sampleid, ps)
metadata_keep <- metadata_keep %>% arrange(match(sampleid, sample_names(ps)))

sample_data(ps) <- sample_data(metadata_keep %>% column_to_rownames("sampleid"))

# Remove zero-count samples and zero-count ASVs.
ps <- prune_samples(sample_sums(ps) > 0, ps)
ps <- prune_taxa(taxa_sums(ps) > 0, ps)

sample_data(ps)$site <- droplevels(sample_data(ps)$site)

ps
```

# Sequencing depth and rarefaction

```{r sequencing-depth}
read_depth <- tibble(
  sampleid = sample_names(ps),
  reads = sample_sums(ps)
) %>%
  left_join(data.frame(sample_data(ps)) %>% rownames_to_column("sampleid"), by = "sampleid")

p_depth <- ggplot(read_depth, aes(x = site, y = reads, fill = site)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.12, size = 2, alpha = 0.85) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Reads per sample", fill = "Site") +
  theme_bw() +
  theme(legend.position = "none")

p_depth

ggsave(file.path(fig_dir, "read_depth_18S.png"), p_depth, width = 5.5, height = 4.5, dpi = 300)
```

```{r rarefaction-curves, fig.width=7, fig.height=5}
p_rarefaction <- ranacapa::ggrare(ps, step = 1000, se = FALSE) +
  facet_wrap(~ site) +
  theme_bw()

p_rarefaction

ggsave(file.path(fig_dir, "rarefaction_curves_18S.png"), p_rarefaction, width = 8, height = 5, dpi = 300)
```

# Alpha diversity

```{r alpha-table}

alpha_df <- estimate_richness(ps, measures = c("Observed", "Shannon")) %>%
  rownames_to_column("sampleid") %>%
  left_join(data.frame(sample_data(ps)) %>% rownames_to_column("sampleid"), by = "sampleid") %>%
  mutate(site = factor(site, levels = c("Control", "High-CO2")))

alpha_summary <- alpha_df %>%
  group_by(site) %>%
  summarise(
    n = n(),
    mean_observed = mean(Observed, na.rm = TRUE),
    sd_observed = sd(Observed, na.rm = TRUE),
    se_observed = sd_observed / sqrt(n),
    mean_shannon = mean(Shannon, na.rm = TRUE),
    sd_shannon = sd(Shannon, na.rm = TRUE),
    se_shannon = sd_shannon / sqrt(n),
    .groups = "drop"
  )

alpha_summary
```

```{r alpha-plots}
plot_alpha <- function(metric, ylab){
  ggplot(alpha_df, aes(x = site, y = .data[[metric]], fill = site)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.12, size = 2, alpha = 0.85) +
    labs(x = NULL, y = ylab, fill = "Site") +
    theme_bw() +
    theme(legend.position = "none")
}

p_observed <- plot_alpha("Observed", "Observed ASVs")
p_shannon  <- plot_alpha("Shannon", "Shannon diversity")

p_observed
p_shannon

ggsave(file.path(fig_dir, "alpha_observed_18S.png"), p_observed, width = 5.5, height = 4.5, dpi = 300)
ggsave(file.path(fig_dir, "alpha_shannon_18S.png"),  p_shannon,  width = 5.5, height = 4.5, dpi = 300)
```

## Alpha diversity models

```{r alpha-models}
mod_observed <- lm(Observed ~ site, data = alpha_df)
mod_shannon  <- lm(Shannon  ~ site, data = alpha_df)

anova(mod_observed)
anova(mod_shannon)

summary(mod_observed)
summary(mod_shannon)
```

# Beta diversity: Bray-Curtis

```{r beta-functions}

ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

make_pcoa_df <- function(ps_obj, distance_object){
  ord <- ordinate(ps_obj, method = "PCoA", distance = distance_object)
  eig <- ord$values$Relative_eig * 100
  scores <- data.frame(ord$vectors[, 1:2]) %>%
    rownames_to_column("sampleid") %>%
    rename(PCoA1 = Axis.1, PCoA2 = Axis.2) %>%
    left_join(data.frame(sample_data(ps_obj)) %>% rownames_to_column("sampleid"), by = "sampleid")
  list(scores = scores, eig = eig, ord = ord)
}

plot_pcoa <- function(pcoa, colour_var = "site", shape_var = NULL, title = NULL){
  aes_base <- aes(x = PCoA1, y = PCoA2, colour = .data[[colour_var]])
  if(!is.null(shape_var)){
    aes_base <- aes(x = PCoA1, y = PCoA2, colour = .data[[colour_var]], shape = .data[[shape_var]])
  }

  ggplot(pcoa$scores, aes_base) +
    geom_point(size = 3, alpha = 0.9) +
    stat_ellipse(aes(group = .data[[colour_var]]), linewidth = 0.5, show.legend = FALSE) +
    labs(
      x = paste0("PCoA1 (", round(pcoa$eig[1], 1), "%)"),
      y = paste0("PCoA2 (", round(pcoa$eig[2], 1), "%)"),
      colour = colour_var,
      shape = shape_var,
      title = title
    ) +
    theme_bw()
}
```

```{r beta-bray}

bray_dist <- phyloseq::distance(ps_rel, method = "bray")
bray_pcoa <- make_pcoa_df(ps_rel, bray_dist)

p_bray <- plot_pcoa(
  bray_pcoa,
  colour_var = "site",
  title = "18S beta diversity: Bray-Curtis"
)

p_bray

ggsave(file.path(fig_dir, "beta_bray_18S.png"), p_bray, width = 6.5, height = 5, dpi = 300)
```

## PERMANOVA and beta dispersion

```{r permanova-dispersion}
meta_beta <- data.frame(sample_data(ps_rel), check.names = FALSE)
meta_beta <- meta_beta[labels(bray_dist), , drop = FALSE]

adonis2(bray_dist ~ site, data = meta_beta, permutations = 9999)

bd_site <- betadisper(bray_dist, group = meta_beta$site)
anova(bd_site)
permutest(bd_site, permutations = 9999)
plot(bd_site)
```

# Taxa-composition barplots

This section plots the most abundant taxa at a selected taxonomic rank. The default rank is `Phylum`, but you can change this to `Kingdom`, `Class`, `Order`, `Family`, or `Genus` depending on the quality of the 18S taxonomy.

```{r taxa-barplot-functions}
barplot_rank <- "Phylum"
top_n_taxa <- 12

available_ranks <- rank_names(ps)
available_ranks

if(!barplot_rank %in% available_ranks){
  stop("barplot_rank is not present in rank_names(ps). Available ranks are: ", paste(available_ranks, collapse = ", "))
}

make_taxa_barplot_data <- function(ps_obj, rank = "Phylum", top_n = 12){
  ps_rank <- tax_glom(ps_obj, taxrank = rank, NArm = FALSE)
  ps_rank_rel <- transform_sample_counts(ps_rank, function(x) x / sum(x))

  taxa_df <- psmelt(ps_rank_rel) %>%
    mutate(
      taxon = as.character(.data[[rank]]),
      taxon = if_else(is.na(taxon) | taxon == "" | taxon == "NA", "Unclassified", taxon)
    )

  top_taxa <- taxa_df %>%
    group_by(taxon) %>%
    summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abundance)) %>%
    slice_head(n = top_n) %>%
    pull(taxon)

  taxa_df %>%
    mutate(taxon = if_else(taxon %in% top_taxa, taxon, "Other")) %>%
    group_by(sampleid = Sample, site, taxon) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop") %>%
    mutate(taxon = fct_reorder(taxon, Abundance, .fun = sum, .desc = TRUE))
}

taxa_bar_df <- make_taxa_barplot_data(ps, rank = barplot_rank, top_n = top_n_taxa)
```

## Per-sample taxa barplot

```{r taxa-barplot-samples, fig.width=10, fig.height=6}
p_taxa_samples <- ggplot(taxa_bar_df, aes(x = sampleid, y = Abundance, fill = taxon)) +
  geom_col(width = 0.9) +
  facet_grid(~ site, scales = "free_x", space = "free_x") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Sample", y = "Relative abundance", fill = barplot_rank) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    panel.spacing.x = unit(0.8, "lines")
  )

p_taxa_samples

ggsave(file.path(fig_dir, paste0("taxa_barplot_samples_", barplot_rank, "_18S.png")), p_taxa_samples, width = 10, height = 6, dpi = 300)
```

## Mean taxa barplot by site

```{r taxa-barplot-mean, fig.width=6, fig.height=5}
taxa_mean_df <- taxa_bar_df %>%
  group_by(site, taxon) %>%
  summarise(mean_abundance = mean(Abundance, na.rm = TRUE), .groups = "drop")

p_taxa_mean <- ggplot(taxa_mean_df, aes(x = site, y = mean_abundance, fill = taxon)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Mean relative abundance", fill = barplot_rank) +
  theme_bw()

p_taxa_mean

ggsave(file.path(fig_dir, paste0("taxa_barplot_mean_site_", barplot_rank, "_18S.png")), p_taxa_mean, width = 6.5, height = 5, dpi = 300)
```

# Optional UniFrac analyses

Only run this section if the rooted tree is present and reliable.

```{r optional-unifrac, eval=FALSE}
unifrac_dist <- UniFrac(ps, weighted = FALSE, normalized = TRUE, parallel = FALSE, fast = TRUE)
wunifrac_dist <- UniFrac(ps, weighted = TRUE, normalized = TRUE, parallel = FALSE, fast = TRUE)

unifrac_pcoa <- make_pcoa_df(ps, unifrac_dist)
wunifrac_pcoa <- make_pcoa_df(ps, wunifrac_dist)

p_unifrac <- plot_pcoa(unifrac_pcoa, colour_var = "site", title = "18S beta diversity: unweighted UniFrac")
p_wunifrac <- plot_pcoa(wunifrac_pcoa, colour_var = "site", title = "18S beta diversity: weighted UniFrac")

p_unifrac
p_wunifrac

ggsave(file.path(fig_dir, "beta_unweighted_unifrac_18S.png"), p_unifrac, width = 6.5, height = 5, dpi = 300)
ggsave(file.path(fig_dir, "beta_weighted_unifrac_18S.png"),   p_wunifrac, width = 6.5, height = 5, dpi = 300)

meta_unifrac <- data.frame(sample_data(ps), check.names = FALSE)
meta_unifrac <- meta_unifrac[labels(unifrac_dist), , drop = FALSE]

adonis2(unifrac_dist ~ site, data = meta_unifrac, permutations = 9999)
adonis2(wunifrac_dist ~ site, data = meta_unifrac, permutations = 9999)
```

```{r}

plot_krona(ps, "krona_output.html", variable = "site")

```


# Save summary tables

```{r save-tables}
write_csv(alpha_df, file.path(fig_dir, "alpha_diversity_sample_values_18S.csv"))
write_csv(alpha_summary, file.path(fig_dir, "alpha_diversity_summary_18S.csv"))
write_csv(bray_pcoa$scores, file.path(fig_dir, "bray_pcoa_scores_18S.csv"))
write_csv(taxa_bar_df, file.path(fig_dir, paste0("taxa_barplot_sample_values_", barplot_rank, "_18S.csv")))
write_csv(taxa_mean_df, file.path(fig_dir, paste0("taxa_barplot_mean_site_", barplot_rank, "_18S.csv")))
```
