---
title: "PostKrakenBacterialMetatranscriptome"
author: "Callum Hudson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Diamond taxonomy load
```{r}
library(data.table)
library(dplyr)

# DIAMOND
diamond <- fread(
  "PostKraken2/transcripts_gtdb.tsv",
  header = FALSE
)

colnames(diamond) <- c(
  "ORF",
  "subject",
  "taxid",
  "pident",
  "aln_length",
  "evalue",
  "bitscore"
)

# GTDB nodes
nodes <- fread(
  "DIAMOND_tax_R_files/gtdb-r220.nodes.dmp",
  sep="|",
  header=FALSE,
  fill=TRUE,
  strip.white=TRUE
)

nodes <- nodes[, .(
  taxid = V1,
  parent = V2,
  rank = V3
)]

# GTDB names
names <- fread(
  "DIAMOND_tax_R_files/gtdb-r220.names.dmp",
  sep="|",
  header=FALSE,
  fill=TRUE,
  strip.white=TRUE
)

names <- names[, .(
  taxid = V1,
  name = V2
)]

nodes$rank <- trimws(nodes$rank)

names$name <- trimws(names$name)

head(nodes)
head(names)

sort(unique(nodes$rank))
length(unique(diamond$taxid))
```

# Make taxonomy lookup table 
```{r}
# ------------------------------------------------------------------
# Build lookup vectors
# ------------------------------------------------------------------

parent_lookup <- setNames(nodes$parent, nodes$taxid)
rank_lookup   <- setNames(nodes$rank, nodes$taxid)
name_lookup   <- setNames(names$name, names$taxid)

# ------------------------------------------------------------------
# Function to recover full lineage for one taxid
# ------------------------------------------------------------------

get_lineage <- function(taxid){

  lineage <- c(
    superkingdom = NA,
    phylum = NA,
    class = NA,
    order = NA,
    family = NA,
    genus = NA,
    species = NA
  )

  current <- as.character(taxid)

  visited <- character()

  while(!is.na(current) &&
        current %in% names(parent_lookup) &&
        !(current %in% visited)){

    visited <- c(visited, current)

    rank <- rank_lookup[current]
    name <- name_lookup[current]

    if(rank %in% names(lineage)){
      lineage[rank] <- name
    }

    parent <- parent_lookup[current]

    if(is.na(parent) || parent == current){
      break
    }

    current <- as.character(parent)
  }

  data.frame(
    taxid = as.character(taxid),
    t(lineage),
    stringsAsFactors = FALSE
  )
}

# ------------------------------------------------------------------
# Build taxonomy table for all DIAMOND taxids
# ------------------------------------------------------------------

unique_taxids <- unique(as.character(diamond$taxid))

taxonomy_table <- rbindlist(
  lapply(unique_taxids, get_lineage),
  fill = TRUE
)

# ------------------------------------------------------------------
# Inspect
# ------------------------------------------------------------------

dim(taxonomy_table)

head(taxonomy_table)

table(taxonomy_table$superkingdom)
```

```{r}
fwrite(
  taxonomy_table,
  "PostKraken2/diamond_taxonomy_table.tsv",
  sep = "\t"
)

taxonomy_table <- fread("PostKraken2/diamond_taxonomy_table.tsv")
```

```{r}
diamond$taxid <- as.character(diamond$taxid)
taxonomy_table$taxid <- as.character(taxonomy_table$taxid)

diamond_tax <- left_join(
  diamond,
  taxonomy_table,
  by = "taxid"
) 

View(diamond_tax)
```

# Load salmon results 
# No need to run again, just load the RDS in the chunk below
```{r}
library(data.table)

# ------------------------------------------------------------------
# Find Salmon files
# ------------------------------------------------------------------

files <- list.files(
  "PostKraken2/Read_mapping_salmon",
  pattern = "_quant\\.sf$",
  full.names = TRUE
)

# ------------------------------------------------------------------
# Pass 1: count samples with >0 reads within each treatment
# ------------------------------------------------------------------

counts <- NULL

for(i in seq_along(files)) {

  sample_name <- basename(files[i])

  cat("Counting:", sample_name, "\n")

  dt <- fread(
    files[i],
    select = c("Name", "NumReads")
  )

  detected <- as.integer(dt$NumReads > 1)

  if(i == 1) {

    counts <- data.table(
      ORF = dt$Name,
      C = 0L,
      H = 0L
    )

  }

  if(grepl("^C", sample_name)) {

    counts$C <- counts$C + detected

  } else if(grepl("^H", sample_name)) {

    counts$H <- counts$H + detected

  } else {

    stop("Unknown sample name: ", sample_name)

  }

  rm(dt, detected)
  gc()
}

# ------------------------------------------------------------------
# Keep ORFs with >0 reads in >=7 samples of either treatment
# ------------------------------------------------------------------

keep_orfs <- counts[
  C >= 7 | H >= 7,
  ORF
]

cat("ORFs retained:", length(keep_orfs), "\n")

# Counts table no longer needed
rm(counts)
gc()

# ------------------------------------------------------------------
# Pass 2: load only retained ORFs
# ------------------------------------------------------------------

salmon_list <- vector("list", length(files))

for(i in seq_along(files)) {

  cat("Loading:", basename(files[i]), "\n")

  sample_name <- sub(
    "_quant\\.sf$",
    "",
    basename(files[i])
  )

  dt <- fread(
    files[i],
    select = c("Name", "TPM", "NumReads")
  )

  dt <- dt[Name %in% keep_orfs]

  setnames(dt, "Name", "ORF")

  dt[, Sample := sample_name]

  salmon_list[[i]] <- dt

  rm(dt)
  gc()
}

salmon_filtered <- rbindlist(salmon_list)

# ------------------------------------------------------------------
# Check result
# ------------------------------------------------------------------

print(dim(salmon_filtered))
head(salmon_filtered)

# ------------------------------------------------------------------
# Save filtered object
# ------------------------------------------------------------------

saveRDS(
  salmon_filtered,
  "PostKraken2/00_TempData/salmon_filtered.rds",
  compress = "xz"
)
```

```{r}
salmon_filtered <- readRDS("PostKraken2/00_TempData/salmon_filtered.rds")
```

```{r}
library(data.table)
library(dplyr)
library(tidyr)
library(compositions)
library(ggplot2)

# =========================================================
# 1. Species-level abundance table
# =========================================================

bac_sp <- bac %>%
  mutate(species = ifelse(is.na(species), "Unassigned", species)) %>%
  group_by(Sample, Site, species) %>%
  summarise(reads = sum(NumReads, na.rm = TRUE), .groups = "drop")

sp_mat <- bac_sp %>%
  pivot_wider(
    names_from = species,
    values_from = reads,
    values_fill = 0
  )

# =========================================================
# 2. Build matrix + metadata (SAFE ALIGNMENT)
# =========================================================

meta <- sp_mat %>%
  dplyr::select(Sample, Site) %>%
  tibble::column_to_rownames("Sample")

mat <- sp_mat %>%
  dplyr::select(-Sample, -Site)

mat <- as.matrix(mat)

rownames(mat) <- sp_mat$Sample

# ensure numeric (IMPORTANT safety step)
storage.mode(mat) <- "numeric"

# align metadata
meta <- meta[rownames(mat), , drop = FALSE]

stopifnot(all(rownames(meta) == rownames(mat)))

# =========================================================
# 3. CLR transform
# =========================================================

mat_clr <- clr(mat + 1)

# =========================================================
# 4. PCA (CLR space)
# =========================================================

pca <- prcomp(mat_clr, scale. = FALSE)

percentVar <- round(100 * (pca$sdev^2 / sum(pca$sdev^2)), 1)

scores <- as.data.frame(pca$x[, 1:2])
scores$Sample <- rownames(scores)
scores <- cbind(scores, meta)

# =========================================================
# 5. Ellipses (PC1 vs PC2)
# =========================================================

veganCovEllipse <- function(cov, center = c(0, 0), npoints = 100) {
  theta <- seq(0, 2 * pi, length.out = npoints)
  circle <- cbind(cos(theta), sin(theta))
  t(center + t(circle %*% chol(cov)))
}

scores$Site <- factor(scores$Site)

ell <- do.call(rbind, lapply(levels(scores$Site), function(g) {

  grp <- scores[scores$Site == g, c("PC1", "PC2")]

  if (nrow(grp) < 3) return(NULL)  # slightly stricter for stability

  cov_data <- cov.wt(grp)$cov
  center <- colMeans(grp)

  e <- veganCovEllipse(cov_data, center)

  data.frame(
    PC1 = e[,1],
    PC2 = e[,2],
    Site = g
  )
}))

# =========================================================
# 6. Plot (PC1 vs PC2 only)
# =========================================================

ggplot(scores, aes(PC1, PC2, color = Site)) +
  geom_point(size = 1.5) +
  geom_path(
    data = ell,
    aes(PC1, PC2, color = Site),
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  scale_color_manual(values = c(
    "Reference" = "#2C728E",
    "Elevated CO2" = "#D4476C"
  )) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    x = paste0("PC1 (", percentVar[1], "% variance)"),
    y = paste0("PC2 (", percentVar[2], "% variance)"),
    title = "Species-level CLR PCA"
  )
```

# Expression analysis (DESEQ2)
```{r}
library(tidyr)
library(DESeq2)

length(unique(salmon_filtered$ORF))

counts_mat <- salmon_filtered %>%
  dplyr::select(ORF, Sample, NumReads) %>%
  pivot_wider(
    names_from = Sample,
    values_from = NumReads,
    values_fill = 0
  )

counts_df <- as.data.frame(counts_mat)

rownames(counts_df) <- counts_df$ORF

counts_df$ORF <- NULL

counts_mat_raw <- as.matrix(counts_df)
dim(counts_mat_raw)

storage.mode(counts_mat_raw) <- "numeric"

metadata <- data.frame(
  Site = c(
    rep("Reference", 14),
    rep("Elevated CO2", 17)
  ),
  row.names = colnames(counts_mat_raw)
)

dim(counts_mat_raw)

colnames(counts_mat_raw)

# Altered filtered low counts, so 
dds <- DESeqDataSetFromMatrix(
  countData = round(counts_mat_raw),
  colData = metadata,
  design = ~ Site
)

metadata <- as.data.frame(colData(dds))

# Run DESeq2
dds <- DESeq(dds)

# VST
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)

```

# Load in so don't need to re-run deseq
```{r}
saveRDS(
  dds,
  file = "PostKraken2/00_TempData/microbial_dds.rds"
)

saveRDS(
  vsd,
  file = "PostKraken2/00_TempData/microbial_vsd.rds"
)

dds <- readRDS("PostKraken2/00_TempData/microbial_dds.rds")

vsd <- readRDS("PostKraken2/00_TempData/microbial_vsd.rds")
```

# Site-wise PCA (all transcripts)
```{r}
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(patchwork)

# =========================================================
# Extract VST matrix and metadata
# =========================================================

vst_mat <- assay(vsd)
metadata <- as.data.frame(colData(dds))

stopifnot(identical(colnames(vst_mat), rownames(metadata)))

# =========================================================
# PCA plotting function
# =========================================================

plot_pca_ellipses <- function(
  vst_mat,
  sample_meta,
  title_text,
  colors = c(
    "Reference" = "#2C728E",
    "Elevated CO2" = "#D4476C"
  )
){

  sample_meta$Site <- factor(
    sample_meta$Site,
    levels = c("Reference", "Elevated CO2")
  )

  stopifnot(identical(colnames(vst_mat),
                      rownames(sample_meta)))

  # PCA on samples
  expr_t <- t(vst_mat)

  pca_res <- prcomp(
    expr_t,
    scale. = FALSE
  )

  percentVar <- round(
    100 * (pca_res$sdev^2 / sum(pca_res$sdev^2)),
    1
  )

  pca_scores <- as.data.frame(
    pca_res$x[,1:3]
  )

  colnames(pca_scores) <- c("PC1","PC2","PC3")

  pca_scores$Sample <- rownames(pca_scores)

  pca_scores <- cbind(
    pca_scores,
    sample_meta[rownames(pca_scores), , drop = FALSE]
  )

  # -------------------------------------------------------
  # Ellipse helper
  # -------------------------------------------------------

  veganCovEllipse <- function(cov,
                              center = c(0,0),
                              npoints = 100){

    theta <- seq(0, 2*pi, length.out = npoints)

    circle <- cbind(
      cos(theta),
      sin(theta)
    )

    t(
      center +
        t(circle %*% chol(cov))
    )
  }

  make_ellipse <- function(x, y){

    do.call(rbind,
      lapply(levels(pca_scores$Site), function(g){

        grp <- pca_scores[
          pca_scores$Site == g,
          c(x,y)
        ]

        if(nrow(grp) < 3)
          return(NULL)

        if(any(!is.finite(as.matrix(grp))))
          return(NULL)

        cov_data <- cov(grp)

        if(det(cov_data) <= 0)
          return(NULL)

        center <- colMeans(grp)

        ell <- veganCovEllipse(
          cov_data,
          center
        )

        data.frame(
          x = ell[,1],
          y = ell[,2],
          Site = g
        )

      })
    )

  }

  ell12 <- make_ellipse("PC1","PC2")
  ell13 <- make_ellipse("PC1","PC3")
  ell23 <- make_ellipse("PC2","PC3")

  # -------------------------------------------------------
  # Plot helper
  # -------------------------------------------------------

  base_plot <- function(x,y,ell,xlab,ylab){

    ggplot(
      pca_scores,
      aes_string(
        x = x,
        y = y,
        colour = "Site"
      )
    ) +

      geom_point(size = 2) +

      geom_path(
        data = ell,
        aes(
          x = x,
          y = y,
          colour = Site
        ),
        linewidth = 0.8,
        inherit.aes = FALSE
      ) +

      scale_color_manual(values = colors) +

      labs(
        x = xlab,
        y = ylab,
        title = title_text
      ) +

      theme_bw() +

      theme(
        panel.grid = element_blank(),
        legend.position = "bottom"
      )

  }

  p1 <- base_plot(
    "PC1",
    "PC2",
    ell12,
    paste0("PC1 (",percentVar[1],"%)"),
    paste0("PC2 (",percentVar[2],"%)")
  ) 

  p2 <- base_plot(
    "PC1",
    "PC3",
    ell13,
    paste0("PC1 (",percentVar[1],"%)"),
    paste0("PC3 (",percentVar[3],"%)")
  )

  p3 <- base_plot(
    "PC2",
    "PC3",
    ell23,
    paste0("PC2 (",percentVar[2],"%)"),
    paste0("PC3 (",percentVar[3],"%)")
  )

  (p1 + p2 + p3) +
    plot_annotation(
      tag_levels = "A"
    ) +
    plot_layout(
      guides = "collect"
    ) &
    theme(
      legend.position = "right",
      plot.tag = element_text(
        face = "bold",
        size = 14
      )
    )

}

# =========================================================
# Create PCA
# =========================================================

PCA_microbes <- plot_pca_ellipses(
  vst_mat = vst_mat,
  sample_meta = metadata,
  title_text = ""
) 

PCA_microbes

# =========================================================
# Save figure
# =========================================================

ggsave(
  "PostKraken2/Figures/Microbial_PCA.svg",
  PCA_microbes,
  width = 10,
  height = 4
)
```

# Statistical test (Reference vs Elevated CO2 expression pattern - all ORFs)
```{r}
nrow(dds)

library(vegan)

# samples x genes
expr_t <- t(vst_mat)

# Euclidean distance on VST values
dist_mat <- dist(expr_t, method = "euclidean")

# PERMANOVA
adonis_all <- adonis2(
  dist_mat ~ Site,
  data = metadata,
  permutations = 999,
  seed = 1234
)

adonis_all

#adonis2(formula = dist_mat ~ Site, data = metadata, permutations = 999, seed = 1234)
#         Df SumOfSqs     R2      F Pr(>F)  
#Model     1    14879 0.0779 2.4501  0.017 *
#Residual 29   176117 0.9221                
#Total    30   190996 1.0000     

# Test homogeneity of dispersion
bd_all <- betadisper(
  dist_mat,
  metadata$Site
)

anova(bd_all)

#Response: Distances
#          Df Sum Sq Mean Sq F value  Pr(>F)  
#Groups     1 1228.7 1228.70  4.4697 0.04322 *
#Residuals 29 7971.9  274.89  

permutest(
  bd_all,
  permutations = 999,
  seed = 1234
)

#Response: Distances
#          Df Sum Sq Mean Sq      F N.Perm Pr(>F)  
#Groups     1 1228.7 1228.70 4.4697    999  0.043 *
#Residuals 29 7971.9  274.89   
```


# Bacterial assigned ORFs only
```{r}
library(dplyr)
library(tidyr)
library(DESeq2)

# =========================================================
# Build count matrix from bacterial ORFs only
# =========================================================

counts_mat <- bac %>%
  dplyr::select(ORF, Sample, NumReads) %>%
  pivot_wider(
    names_from = Sample,
    values_from = NumReads,
    values_fill = 0
  )

counts_df <- as.data.frame(counts_mat)

rownames(counts_df) <- counts_df$ORF
counts_df$ORF <- NULL

counts_mat_raw <- as.matrix(counts_df)

storage.mode(counts_mat_raw) <- "numeric"

# =========================================================
# Metadata
# =========================================================

metadata <- data.frame(
  Site = c(
    rep("Reference", 14),
    rep("Elevated CO2", 17)
  ),
  row.names = colnames(counts_mat_raw)
)

# =========================================================
# DESeq2 dataset
# =========================================================

dds_bac <- DESeqDataSetFromMatrix(
  countData = round(counts_mat_raw),
  colData = metadata,
  design = ~ Site
)

# Remove genes with zero counts
dds_bac <- dds_bac[rowSums(counts(dds_bac)) > 0, ]

# =========================================================
# Run DESeq2
# =========================================================

dds_bac  <- DESeq(dds_bac)

# =========================================================
# Variance stabilizing transformation
# =========================================================

vsd_bac <- varianceStabilizingTransformation(
  dds_bac ,
  blind = FALSE
)

# =========================================================
# Save objects
# =========================================================

saveRDS(
  dds_bac ,
  file = "PostKraken2/00_TempData/bacterial_only_dds.rds"
)

saveRDS(
  vsd_bac,
  file = "PostKraken2/00_TempData/bacterial_only_vsd.rds"
)

# =========================================================
# PCA
# =========================================================

vst_mat_bac <- assay(vsd_bac)
metadata_bac <- as.data.frame(colData(dds_bac))

PCA_microbes <- plot_pca_ellipses(
    vst_mat = vst_mat_bac,
    sample_meta = metadata_bac,
    title_text = ""
)

PCA_microbes

ggsave(
  "PostKraken2/Figures/Bacterial_ORF_PCA.svg",
  PCA_microbes,
  width = 10,
  height = 4
)
```

# Statistical test (Reference vs Elevated CO2 expression pattern - All ORFs and Bacteria only)
```{r}
nrow(dds_bac)
#### Bacteria only #####
library(vegan)

vst_mat_bac <- assay(vsd_bac)
metadata_bac <- as.data.frame(colData(dds_bac))

# samples x genes
expr_t_bac <- t(vst_mat_bac)

# Euclidean distance on VST values
dist_mat_bac <- dist(expr_t_bac, method = "euclidean")

# PERMANOVA
adonis_bacteria <- adonis2(
  dist_mat_bac ~ Site,
  data = metadata_bac,
  permutations = 999,
  seed = 1234
)

adonis_bacteria

#adonis2(formula = dist_mat_bac ~ Site, data = metadata_bac, permutations = 999, seed = 1234)
#         Df SumOfSqs      R2      F Pr(>F)  
#Model     1    13853 0.07684 2.4137  0.019 *
#Residual 29   166435 0.92316                
#Total    30   180287 1.00000  

# Test homogeneity of dispersion
bd_bacteria <- betadisper(
  dist_mat_bac,
  metadata_bac$Site
)

anova(bd_bacteria)

#Response: Distances
#          Df Sum Sq Mean Sq F value  Pr(>F)  
#Groups     1 1200.6 1200.58  4.7747 0.03711 *
#Residuals 29 7291.9  251.44    

permutest(
  bd_bacteria,
  permutations = 999,
  seed = 1234
)

#Response: Distances
#          Df Sum Sq Mean Sq      F N.Perm Pr(>F)  
#Groups     1 1200.6 1200.58 4.7747    999  0.041 *
#Residuals 29 7291.9  251.44  
```

# KEGG Pathways
```{r}
BiocManager::install(c(
  "clusterProfiler",
  "enrichplot",
  "DOSE"
))

# ============================================================
# 0. LIBRARIES
# ============================================================
library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(DESeq2)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)

# ============================================================
# 1. LOAD SALMON DATA
# ============================================================
#salmon_filtered <- readRDS("00_TempData/salmon_filtered.rds")

expr <- bac %>%
  dplyr::select(ORF, Sample, NumReads)

cat("\nUnique Salmon transcripts:\n")
print(n_distinct(expr$ORF))

# ============================================================
# 2. LOAD eggNOG ANNOTATION
# ============================================================
eggnog <- fread(
  "PostKraken2/postkraken_eggnog.emapper.annotations",
  comment.char = "#",
  header = FALSE,
  fill = TRUE,
  sep = "\t"
)


colnames(eggnog) <- c(
  "query","seed_ortholog","evalue","score","eggNOG_OGs",
  "max_annot_lvl","COG_category","Description","Preferred_name",
  "GOs","EC","KEGG_ko","KEGG_Pathway","KEGG_Module",
  "KEGG_Reaction","KEGG_rclass","BRITE","KEGG_TC","CAZy",
  "BiGG_Reaction","PFAMs"
)

cat("\nUnique eggNOG ORFs:\n")
print(n_distinct(eggnog$query))

# ============================================================
# 3. CONVERT PRODIGAL ORF IDs -> PARENT TRANSCRIPT IDs
# ============================================================
# Example:
# NODE_xxx_i0_1 -> NODE_xxx_i0
# NODE_xxx_i0_2 -> NODE_xxx_i0

eggnog <- eggnog %>%
  mutate(
    ORF = str_remove(query, "_[0-9]+$")
  )

# ============================================================
# 4. KEEP ONLY TRANSCRIPTS PRESENT IN SALMON
# ============================================================
keep_orfs <- unique(expr$ORF)

eggnog <- eggnog %>%
  filter(ORF %in% keep_orfs)

cat("\nAnnotated transcripts overlapping Salmon:\n")
print(n_distinct(eggnog$ORF))

# ============================================================
# 5. CLEAN KEGG KO ANNOTATIONS
# ============================================================
eggnog_clean <- eggnog %>%
  dplyr::select(ORF, KEGG_ko) %>%
  filter(
    !is.na(KEGG_ko),
    KEGG_ko != "-"
  ) %>%
  mutate(
    KEGG_ko = str_replace_all(KEGG_ko, "ko:", "")
  )

# split multiple KO assignments
eggnog_long <- eggnog_clean %>%
  separate_rows(KEGG_ko, sep = ",|;")

cat("\nUnique KEGG KOs:\n")
print(n_distinct(eggnog_long$KEGG_ko))

# ============================================================
# 6. MERGE EXPRESSION + FUNCTION
# ============================================================
expr_fun <- expr %>%
  inner_join(
    eggnog_long,
    by = "ORF"
  )

cat("\nTranscripts after merge:\n")
print(n_distinct(expr_fun$ORF))

cat("\nRows after merge:\n")
print(nrow(expr_fun))

# ============================================================
# 7. BUILD KO COUNT MATRIX
# ============================================================
# Transcript abundance is propagated to all ORFs/KOs
# predicted on that transcript.

kegg_long <- expr_fun %>%
  group_by(Sample, KEGG_ko) %>%
  summarise(
    abundance = sum(NumReads, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nNumber of KEGG KOs in abundance table:\n")
print(n_distinct(kegg_long$KEGG_ko))

kegg_mat <- kegg_long %>%
  pivot_wider(
    names_from = Sample,
    values_from = abundance,
    values_fill = 0
  )

kegg_mat_df <- as.data.frame(kegg_mat)

rownames(kegg_mat_df) <- kegg_mat_df$KEGG_ko
kegg_mat_df$KEGG_ko <- NULL

kegg_mat_df <- as.matrix(kegg_mat_df)

# remove empty KOs
kegg_mat_df <- kegg_mat_df[rowSums(kegg_mat_df) > 0, ]

storage.mode(kegg_mat_df) <- "integer"

cat("\nKO matrix dimensions:\n")
print(dim(kegg_mat_df))

# ============================================================
# 8. SAMPLE METADATA
# ============================================================
# C samples = Reference
# H samples = Elevated CO2

metadata <- data.frame(
  Site = ifelse(
    grepl("^C", colnames(kegg_mat_df)),
    "Reference",
    "Elevated_CO2"
  ),
  row.names = colnames(kegg_mat_df)
)

metadata$Site <- factor(
  metadata$Site,
  levels = c("Reference", "Elevated_CO2")
)

cat("\nMetadata table:\n")
print(table(metadata$Site))

# ============================================================
# 11. KEGG GSEA
# ============================================================

# Build ranked KO list
geneList <- res_df %>%
  filter(
    !is.na(log2FoldChange)
  ) %>%
  distinct(KO, .keep_all = TRUE) %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(KO, log2FoldChange)

geneList <- geneList$log2FoldChange
names(geneList) <- res_df %>%
  filter(
    !is.na(log2FoldChange)
  ) %>%
  distinct(KO, .keep_all = TRUE) %>%
  arrange(desc(log2FoldChange)) %>%
  pull(KO)

geneList <- sort(geneList, decreasing = TRUE)

cat("\nKOs in ranked list:\n")
print(length(geneList))

# ------------------------------------------------------------
# GSEA
# ------------------------------------------------------------

gsea_kegg <- gseKEGG(
  geneList = geneList,
  organism = "ko",
  keyType = "kegg",
  pAdjustMethod = "BH",
  minGSSize = 5,
  maxGSSize = 500,
  pvalueCutoff = 0.05,
  verbose = FALSE, 
  seed = 1234
)

# ------------------------------------------------------------
# Results
# ------------------------------------------------------------

if (!is.null(gsea_kegg) &&
    nrow(as.data.frame(gsea_kegg)) > 0) {

  gsea_df <- as.data.frame(gsea_kegg)

  print(
    gsea_df %>%
      dplyr::select(
        ID,
        Description,
        NES,
        p.adjust
      ) %>%
      arrange(p.adjust) %>%
      head(20)
  )

  # Dotplot
  clusterProfiler::dotplot(
    gsea_kegg,
    showCategory = 20
  )

  # Enrichment plot for top pathway
  top_pathway <- gsea_df$ID[1]

  gseaplot2(
    gsea_kegg,
    geneSetID = top_pathway,
    title = gsea_df$Description[1]
  )

} else {

  message(
    "No significant KEGG pathways detected by GSEA."
  )

}

# ============================================================
# 12. Dot Plot
# ============================================================

library(patchwork)

gsea_pos <- gsea_kegg
gsea_pos@result <- gsea_pos@result %>%
  filter(NES > 0) %>%
  arrange(p.adjust)

gsea_neg <- gsea_kegg
gsea_neg@result <- gsea_neg@result %>%
  filter(NES < 0) %>%
  arrange(p.adjust)

p1 <- clusterProfiler::dotplot(
  gsea_pos,
  showCategory = 20
) +
ggtitle("Enriched") +
theme(
  axis.text.y = element_text(size = 8)
)

p2 <-  clusterProfiler::dotplot(
  gsea_neg,
  showCategory = 20
) +
ggtitle("Suppressed") +
theme(
  axis.text.y = element_text(size = 8)
)

p1 + p2

sum(res_df$padj < 0.05, na.rm = TRUE)

View(plot_df)
```

```{r}
# ============================================================
# Filter KEGG results to bacterial-relevant pathways
# ============================================================

library(KEGGREST)
library(stringr)
library(dplyr)

# Update pathway names from KEGG
pathways <- keggList("pathway", "ko")

pathway_tbl <- data.frame(
  ID = sub("^path:", "", names(pathways)),
  Description = unname(pathways),
  stringsAsFactors = FALSE
)

gsea_kegg@result <- gsea_kegg@result %>%
  left_join(pathway_tbl, by = "ID") %>%
  mutate(
    Description = coalesce(Description.y, Description.x)
  ) %>%
  dplyr::select(-Description.x, -Description.y)

# ------------------------------------------------------------
# Remove pathways that are not informative for bacteria
# ------------------------------------------------------------

remove_patterns <- c(

  # Human disease
  "disease",
  "cancer",
  "carcinogenesis",
  "viral",
  "bacterial infection",
  "parasite",
  "Huntington",
  "Parkinson",
  "Alzheimer",
  "Amyotrophic",
  "ALS",
  "Prion",
  "Glioma",
  "Leukemia",
  "Melanoma",
  "Tuberculosis",
  "COVID",
  "Influenza",

  # Organismal systems
  "immune",
  "endocrine",
  "digestive",
  "circulatory",
  "nervous",
  "reproductive",
  "development",
  "longevity",

  # Animal signalling
  "MAPK",
  "ErbB",
  "Wnt",
  "Notch",
  "Hippo",
  "TGF-beta",
  "VEGF",
  "mTOR",
  "PI3K",
  "Ras",
  "Rap1",
  "cAMP",
  "cGMP",
  "Calcium signaling",

  # Animal physiology
  "synapse",
  "axon",
  "dopaminergic",
  "serotonergic",
  "cholinergic",
  "glutamatergic",
  "GABAergic",
  "insulin",
  "thyroid",
  "estrogen",
  "progesterone",
  "GnRH",
  "oxytocin",
  "cardiac",
  "adrenergic",
  "platelet",
  "Fc gamma",
  "HIF-1",
  "apoptosis",
  "autophagy",
  "endocytosis",
  "phagosome"
)

gsea_kegg@result <- gsea_kegg@result %>%
  filter(
    !str_detect(
      Description,
      regex(
        paste(remove_patterns, collapse = "|"),
        ignore_case = TRUE
      )
    )
  )

# ============================================================
# Split into enriched and suppressed pathways
# ============================================================

gsea_pos <- gsea_kegg
gsea_pos@result <- gsea_pos@result %>%
  filter(NES > 0) %>%
  arrange(p.adjust)

gsea_neg <- gsea_kegg
gsea_neg@result <- gsea_neg@result %>%
  filter(NES < 0) %>%
  arrange(p.adjust)

# ============================================================
# Plot
# ============================================================

library(patchwork)

p1 <- clusterProfiler::dotplot(
  gsea_pos,
  showCategory = 20
) +
  ggtitle("Enriched") +
  theme(axis.text.y = element_text(size = 8))

p2 <- clusterProfiler::dotplot(
  gsea_neg,
  showCategory = 20
) +
  ggtitle("Suppressed") +
  theme(axis.text.y = element_text(size = 8))

p1 + p2
```

# NES barplot
```{r}
# ============================================================
# Combined GSEA bar plot
# ============================================================

library(dplyr)
library(ggplot2)
library(forcats)

gsea_df <- as.data.frame(gsea_kegg)

plot_df <- bind_rows(

  gsea_df %>%
    filter(NES > 0) %>%
    arrange(desc(NES)) %>%
    slice_head(n = 15) %>%
    mutate(direction = "Elevated CO2"),

  gsea_df %>%
    filter(NES < 0) %>%
    arrange(NES) %>%
    slice_head(n = 15) %>%
    mutate(direction = "Reference")

) %>%
  mutate(
    neglog10_p = -log10(p.adjust),
    Description = forcats::fct_reorder(Description, NES)
  )

KEGG_combined_bar_plot <- ggplot(
  plot_df,
  aes(
    x = NES,
    y = Description,
    fill = direction,
    alpha = neglog10_p
  )
) +
  geom_col(
    width = 0.6,
    colour = "black",
    linewidth = 0.5
  ) +
  scale_fill_manual(
    values = c(
      "Reference" = "#2C728E",
      "Elevated CO2" = "#D4476C"
    ),
    name = "Enrichment"
  ) +
  scale_alpha_continuous(
    range = c(0.35, 1),
    name = expression(-log[10]~adjusted~italic(p)~value)
  ) +
  labs(
    x = "Normalized Enrichment Score (NES)",
    y = NULL
  ) +
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(
      colour = "black",
      linewidth = 1,
      fill = NA
    ),
    legend.position = "right",
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 11),
    axis.title.x = element_text(size = 12),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

KEGG_combined_bar_plot

ggsave(
  filename = "PostKraken2/Figures/KEGG_GSEA_NES_combined_filtered.svg",
  plot = KEGG_combined_bar_plot,
  width = 8.5,
  height = 4.5
)
```








