---
title: "Roger_assignments"
author: "Callum Hudson"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(compositions) # Centered Log Ratio Plot
set.seed(123)
```


```{r}
ps_noPNA_nPl <- readRDS(file = "/Users/callumjameshudson/Desktop/Urchin Microbiome/Shikine Urchin Microbiome/ps_noPNA_nPL.rds")

ps_shikine <- prune_samples(
  sample_data(ps_noPNA_nPl)$Saibai != "Saibai" &
  !sample_data(ps_noPNA_nPl)$Type %in% c("Standard", "Blank") &
    sample_data(ps_noPNA_nPl)$Treatment !="V" &
  sample_data(ps_noPNA_nPl)$Location != "Shimoda",
  ps_noPNA_nPl
)

sample_data(ps_shikine)
```

```{r}
# Remove samples (Blanks, +ve controls etc.)
ps_shikine_tissue <- subset_samples(ps_shikine, 
                                    Type2 == "Tissue" & 
                                    sample_names(ps_shikine) != "Shikine_Blank_H2O_S37_R1_001.fastq.gz")

sample_data(ps_shikine_tissue)

view(tax_table(ps_shikine_tissue))

otu_df <- as.data.frame(otu_table(ps_shikine_tissue))

# Remove the column names (i.e., ASV numbers)
colnames(otu_df) <- NULL

# Rename rownames to 1 through 32
rownames(otu_df) <- 1:32

# Export to CSV
write.csv(otu_df, "Y.csv")

sample_data(ps_shikine_tissue)$Treatment <- factor(
  recode(sample_data(ps_shikine_tissue)$Treatment,
         "Control" = "Reference",
         "High" = "Elevated CO2"),
  levels = c("Reference", "Elevated CO2")
)


saveRDS(ps_shikine_tissue, "/Users/callumjameshudson/Desktop/Urchin Microbiome/Shikine Urchin Microbiome/02_Temp_Data/ps_shikine_tissue.rds")

ps_shikine_tissue <- readRDS(file = "/Users/callumjameshudson/Desktop/Urchin Microbiome/Shikine Urchin Microbiome/02_Temp_Data/ps_shikine_tissue.rds")


### ADD phylogenetic tree
GTR.fit <- readRDS(file = "/Users/callumjameshudson/Desktop/Urchin Microbiome/Shikine Urchin Microbiome/02_Temp_Data/fitGTR5.rds")

library(phyloseq)
library(phangorn)

# Check the structure of the tree
is(GTR.fit$tree, "phylo")  # should return TRUE

# Add the tree to your phyloseq object
ps_shikine_tissue <- merge_phyloseq(ps_shikine_tissue, phy_tree(GTR.fit$tree))

ps_shikine_tissue 
ps_plot <- subset_samples(ps_shikine_tissue, !(SampleID %in% c("Shikine_C11_S11", "Shikine_C16_S16", "Shikine_H3_S19")))

```

# Extract Taxa Table for MIMIX
```{r}
# Extract taxonomy matrix
tax_mat <- as.matrix(tax_table(ps_gut))

print(dim(tax_mat))
print(head(tax_mat))

# Define rank prefixes, assuming standard ranks in order:
rank_prefixes <- c("k__", "p__", "c__", "o__", "f__", "g__", "s__")

# Make sure number of ranks matches number of columns
if(length(rank_prefixes) != ncol(tax_mat)) {
  stop("Number of rank prefixes does not match number of taxonomy columns!")
}

# Add prefixes column-wise with a loop
tax_mat_prefixed <- matrix(nrow = nrow(tax_mat), ncol = ncol(tax_mat))
for (i in seq_len(ncol(tax_mat))) {
  tax_mat_prefixed[, i] <- ifelse(
    grepl("^\\w+__", tax_mat[, i]),  # if prefix already there, keep as is
    tax_mat[, i],
    paste0(rank_prefixes[i], tax_mat[, i])
  )
}

rownames(tax_mat_prefixed) <- rownames(tax_mat)
colnames(tax_mat_prefixed) <- colnames(tax_mat)

# Combine each row into a semicolon-separated string
taxa_strings <- apply(tax_mat_prefixed, 1, function(x) paste(x, collapse = "; "))

# Make data frame for export
taxa_df <- data.frame(
  taxon_id = rownames(tax_mat),
  taxonomy = taxa_strings,
  stringsAsFactors = FALSE
)

head(taxa_df)

write.csv(taxa_df, "taxa.csv", row.names = FALSE, quote = FALSE)
```


# Load required packages.
```{r, warning=F, message=F, results='hide'}
sapply(c("phyloseq", "ggplot2", "vegan", "pairwiseAdonis"), require, character.only = TRUE)
```

```{r}
level_order1 <- c('Reference', 'Elevated CO2')
disp.pal1 <- c("#2C728E", "#D4476C")
names(disp.pal1) <- level_order1
```

```{r}
# Quick nmds
ps.norm.tissue <- transform_sample_counts(ps_gut_clean, function(x) x / sum(x))

# Ordination
GP.ord <- ordinate(ps.norm.tissue, method = "NMDS", distance = "bray")

# Plot
nMDS <- plot_ordination(ps.norm.tissue, GP.ord, type = "samples", color = "Condition") +
  scale_color_manual(values = disp.pal1)

nMDS
```
```{r}
tree <- phy_tree(ps_gut)

tree_rooted <- midpoint(tree)

phy_tree(ps_gut) <- tree_rooted
```


# Stats for all metrics 
```{r}
##### BRAY Distance #####
bray_bacteria <- phyloseq::distance(ps_gut, method = "bray")

# make a data frame from the sample_data
bacteria.taxa.df <-  data.frame(sample_data(ps_gut)) 

# Beta dispersion test
permutest(betadisper(bray_bacteria, bacteria.taxa.df$Condition))
### multivariate homogeneity of group dispersion OKAY ###
# p = 0.133

# ## PERMANOVA Adonis test
adonis2(formula = bray_bacteria ~ Condition, data = bacteria.taxa.df)
# F = 7.09, p = 0.001 ***

##### Weighted UniFrac #####
# Calculate weighted UniFrac distance matrix
wunifrac_dist <- phyloseq::distance(ps_gut, method = "wunifrac")

# Create sample data frame
bacteria.taxa.df <- data.frame(sample_data(ps_gut))

# 3. Check homogeneity of dispersion
wunifrac_betadisper <- betadisper(wunifrac_dist, bacteria.taxa.df$Condition)
permutest(wunifrac_betadisper)
# p = 0.976 - proceed to PERMANOVA

# 4. Run PERMANOVA
adonis2(formula = wunifrac_dist ~ Condition, data = bacteria.taxa.df)
# F = 9.01, p < 0.001 ***

##### Unweighted UniFrac #####
#  Calculate unweighted UniFrac distance matrix
unifrac_dist <- phyloseq::distance(ps_gut, method = "unifrac")

#  Create sample data frame
bacteria.taxa.df <- data.frame(sample_data(ps_gut))

#  Check homogeneity of dispersion
unifrac_betadisper <- betadisper(unifrac_dist, bacteria.taxa.df$Condition)
permutest(unifrac_betadisper)
# p = 0.486 - proceed to PERMANOVA

# 4. Run PERMANOVA
adonis2(formula = unifrac_dist ~ Condition, data = bacteria.taxa.df)
# F = 2.90, p < 0.001 ***

##### CLR #####

clr_transform <- function(x){
  x <- x + 1                # pseudocount
  clr(x)
}

# CLR transformed OTU table
ps_clr <- transform_sample_counts(ps_gut, clr_transform)

# Euclidean distance (required for CLR)
clr_dist <- phyloseq::distance(ps_clr, method = "euclidean")

# PERMANOVA
adonis_clr <- adonis2(
  clr_dist ~ Condition,
  data = data.frame(sample_data(ps_clr)),
  permutations = 999
)

adonis_clr
# F = 6.20, p = 0.001 ***

# Beta dispersion test
clr_bd <- betadisper(
  clr_dist,
  sample_data(ps_clr)$Condition
)

permutest(clr_bd)
#p = 0.874

##### PhilR #####
adonis_philr <- adonis2(
  philr_dist ~ Condition,
  data = data.frame(sample_data(ps_gut)),
  permutations = 999
)

adonis_philr
# F = 16.77, p < 0.001 ***

# Beta dispersion test
philr_bd <- betadisper(
  philr_dist,
  sample_data(ps_plot)$Condition
)

permutest(philr_bd)
# p = 0.107

##### CLR - Water vs Gut by Site #####
adonis_clr_all <- adonis2(
  clr_dist_all ~ SampleType + Condition,
  data = meta_all,
  permutations = 999,
  by = "margin"
)

adonis_clr_all
#           Df SumOfSqs      R2      F Pr(>F)    
#SampleType  1    35140 0.18968 9.1814  0.001 ***
#Condition   1    23998 0.12954 6.2702  0.001 ***
#Residual   33   126300 0.68176                  
#Total      35   185255 1.00000  

# -------------------------
# Stats: do water sample communities differ?
# -------------------------

ps_water <- subset_samples(
  ps_shikine,
  SampleType == "H2O"
)

# Remove taxa absent from all water samples
ps_water <- prune_taxa(
  taxa_sums(ps_water) > 0,
  ps_water
)

# CLR transformation
ps_water_clr <- transform_sample_counts(
  ps_water,
  clr_transform
)

# Euclidean distance
water_dist <- phyloseq::distance(
  ps_water_clr,
  method = "euclidean"
)

# Metadata
water_meta <- data.frame(
  sample_data(ps_water_clr)
)

# Condition is already correctly coded
water_meta$Condition <- factor(
  water_meta$Condition,
  levels = c("Reference", "Elevated CO2")
)

# PERMANOVA
adonis2(
  water_dist ~ Condition,
  data = water_meta,
  permutations = 999
)

#         Df SumOfSqs      R2      F Pr(>F)
#Model     1   3817.5 0.21647 1.1051    0.2
#Residual  4  13817.3 0.78353              
#Total     5  17634.8 1.00000 

# Water samples between sites didn't significantly differ.
```



### Bray Curtis
```{r}
# Ordinate using Bray-Curtis on relative abundance
GP.ord <- ordinate(
  transform_sample_counts(ps_gut, function(x) x / sum(x)),
  method = "NMDS",
  distance = "bray"
)

# Extract NMDS stress
stress_val <- round(GP.ord$stress, 3)
stress_label <- paste0("Stress = ", stress_val)


# F = 7.09, p = 0.001 ***
adonis_label <- expression(italic(F) == 7.09 * "," ~ italic(p) < 0.001 ~ "***")

# Extract NMDS site scores for ellipses
mds.scores <- as.data.frame(scores(GP.ord, display = "sites"))
mds.scores$Condition <- factor(
  sample_data(ps_gut)$Condition,
  levels = c("Reference", "Elevated CO2")
)

# Ellipse function
veganCovEllipse <- function(cov, center = c(0,0), scale = 1, npoints = 100){
  theta <- (0:npoints) * 2 * pi / npoints
  Circle <- cbind(cos(theta), sin(theta))
  t(center + scale * t(Circle %*% chol(cov)))
}

# Generate ellipse data
df_ellipse <- do.call(rbind, lapply(levels(mds.scores$Condition), function(g){
  group_data <- mds.scores[mds.scores$Condition == g, ]
  cov_data <- cov.wt(cbind(group_data$NMDS1, group_data$NMDS2),
                     wt = rep(1 / nrow(group_data), nrow(group_data)))$cov
  center <- c(mean(group_data$NMDS1), mean(group_data$NMDS2))
  ellipse_coords <- veganCovEllipse(cov_data, center)
  data.frame(NMDS1 = ellipse_coords[,1],
             NMDS2 = ellipse_coords[,2],
             Condition = g)
}))

# Named color palette
disp.pal1 <- c("Reference" = "#2C728E", "Elevated CO2" = "#D4476C")

# --- NMDS plot ---
Bray_nMDS <- plot_ordination(
  ps_gut, GP.ord, type = "samples", color = "Condition"
) +
  geom_path(
    data = df_ellipse,
    aes(x = NMDS1, y = NMDS2, group = Condition, colour = Condition),
    size = 0.8
  ) +
  scale_color_manual(values = disp.pal1) +
  labs(color = "Site") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(fill = NA, color = "black", linewidth = 1),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +
  annotate("text", x = Inf, y = Inf, label = stress_label,
           hjust = 1.05, vjust = 2.8, size = 4) +
  annotate("text", x = Inf, y = Inf, label = adonis_label,
           hjust = 1.05, vjust = 1.25, size = 4, parse = TRUE)

Bray_nMDS

# Save
ggsave("Figures_Alpha_Beta_div/BrayCurtis_NMDS.png", Bray_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
ggsave("Figures_Alpha_Beta_div/BrayCurtis_NMDS.svg", Bray_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
```

### Weighted UniFrac
```{r}
# -----------------------------
# --- Weighted UniFrac NMDS ---
# -----------------------------
# Use Condition directly
w_unifrac_treatment <- factor(sample_data(ps_gut)$Condition,
                              levels = c("Reference", "Elevated CO2"))

# NMDS ordination
GP.ord.w.unifrac <- ordinate(
  transform_sample_counts(ps_gut, function(x) x / sum(x)),
  method = "NMDS",
  distance = "wunifrac"
)

# Stress
stress_val <- round(GP.ord.w.unifrac$stress, 3)
stress_label <- paste0("Stress = ", stress_val)

# F = 8.94, p = 0.002 **
adonis_label <- expression(italic(F) == 9.01 * "," ~ italic(p) < 0.001 ~ "***")

# NMDS scores
w_unifrac_scores <- as.data.frame(scores(GP.ord.w.unifrac, display = "sites"))
w_unifrac_scores$Condition <- w_unifrac_treatment

length(w_unifrac_treatment)
w_unifrac_treatment

# Ellipse function
veganCovEllipse <- function(cov, center = c(0,0), scale = 1, npoints = 100){
  theta <- (0:npoints) * 2 * pi / npoints
  Circle <- cbind(cos(theta), sin(theta))
  t(center + scale * t(Circle %*% chol(cov)))
}

# Generate ellipse data
df_ellipse_w <- do.call(rbind, lapply(levels(w_unifrac_scores$Condition), function(g){
  group_data <- w_unifrac_scores[w_unifrac_scores$Condition == g, ]
  cov_data <- cov.wt(cbind(group_data$NMDS1, group_data$NMDS2),
                     wt = rep(1 / nrow(group_data), nrow(group_data)))$cov
  center <- c(mean(group_data$NMDS1), mean(group_data$NMDS2))
  ellipse_coords <- veganCovEllipse(cov_data, center)
  data.frame(NMDS1 = ellipse_coords[,1],
             NMDS2 = ellipse_coords[,2],
             Condition = g)
}))

# Plot
w_unifrac_nMDS <- plot_ordination(
  ps_plot, GP.ord.w.unifrac, type = "samples", color = "Condition"
) +
  geom_path(data = df_ellipse_w, aes(x = NMDS1, y = NMDS2, colour = Condition, group = Condition), size = 0.8) +
  scale_color_manual(values = disp.pal1) +
  labs(color = "Condition") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(fill = NA, color = "black", linewidth = 1),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +
  annotate("text", x = Inf, y = Inf, label = stress_label,
           hjust = 1.05, vjust = 2.8, size = 4) +
  annotate("text", x = Inf, y = Inf, label = adonis_label,
           hjust = 1.05, vjust = 1.25, size = 4, parse = TRUE)

w_unifrac_nMDS

# Save
ggsave("Figures_Alpha_Beta_div/W_Unifrac_NMDS.png", w_unifrac_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
ggsave("Figures_Alpha_Beta_div/W_Unifrac_NMDS.svg", w_unifrac_nMDS, width = 15, height = 10, units = "cm", dpi = 900)

```

# Unweighted Unifrac
```{r}
# -----------------------------
# --- Unweighted UniFrac NMDS ---
# -----------------------------
unifrac_treatment <- factor(sample_data(ps_plot)$Condition,
                            levels = c("Reference", "Elevated CO2"))

GP.ord.unifrac <- ordinate(
  transform_sample_counts(ps_gut, function(x) x / sum(x)),
  method = "NMDS",
  distance = "unifrac"
)

stress_val <- round(GP.ord.unifrac$stress, 3)
stress_label <- paste0("Stress = ", stress_val)

# F = 2.90, p < 0.001 ***
adonis_label <- expression(italic(F) == 2.90 * "," ~ italic(p) < 0.001 ~ "***")

unifrac_scores <- as.data.frame(scores(GP.ord.unifrac, display = "sites"))
unifrac_scores$Condition <- unifrac_treatment

df_ellipse_u <- do.call(rbind, lapply(levels(unifrac_scores$Condition), function(g){
  group_data <- unifrac_scores[unifrac_scores$Condition == g, ]
  cov_data <- cov.wt(cbind(group_data$NMDS1, group_data$NMDS2),
                     wt = rep(1 / nrow(group_data), nrow(group_data)))$cov
  center <- c(mean(group_data$NMDS1), mean(group_data$NMDS2))
  ellipse_coords <- veganCovEllipse(cov_data, center)
  data.frame(NMDS1 = ellipse_coords[,1],
             NMDS2 = ellipse_coords[,2],
             Condition = g)
}))

UW_unifrac_nMDS <- plot_ordination(
  ps_plot, GP.ord.unifrac, type = "samples", color = "Condition"
) +
  geom_path(data = df_ellipse_u, aes(x = NMDS1, y = NMDS2, colour = Condition, group = Condition), size = 0.8) +
  scale_color_manual(values = disp.pal1) +
  labs(color = "Condition") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(fill = NA, color = "black", linewidth = 1),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +
  annotate("text", x = Inf, y = Inf, label = stress_label,
           hjust = 1.05, vjust = 2.8, size = 4) +
  annotate("text", x = Inf, y = Inf, label = adonis_label,
           hjust = 1.05, vjust = 1.25, size = 4, parse = TRUE)

UW_unifrac_nMDS

# Save
ggsave("Figures_Alpha_Beta_div/UW_Unifrac_NMDS.png", UW_unifrac_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
ggsave("Figures_Alpha_Beta_div/UW_Unifrac_NMDS.svg", UW_unifrac_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
```

# Centered Log Ratio
```{r}
clr_transform <- function(x){
  x <- x + 1                # pseudocount
  clr(x)
}

# Ordinate
GP.ord.clr <- ordinate(
  transform_sample_counts(ps_gut, clr_transform),
  method = "NMDS",
  distance = "euclidean"
)

stress_val <- round(GP.ord.clr$stress, 3)
stress_label <- paste0("Stress = ", stress_val)

# CLR transformed OTU table
ps_clr <- transform_sample_counts(ps_gut, clr_transform)

# Euclidean distance (required for CLR)
clr_dist <- phyloseq::distance(ps_clr, method = "euclidean")

# PERMANOVA
adonis_clr <- adonis2(
  clr_dist ~ Condition,
  data = data.frame(sample_data(ps_clr)),
  permutations = 9999
)

adonis_clr

adonis_label <- expression(italic(F) == 6.20 * "," ~ italic(p) < 0.001 ~ "***")


clr_scores <- as.data.frame(scores(GP.ord.clr, display = "sites"))
clr_scores$Condition <- factor(sample_data(ps_gut)$Condition,
                               levels = c("Reference", "Elevated CO2"))

df_ellipse_clr <- do.call(rbind, lapply(levels(clr_scores$Condition), function(g){

  group_data <- clr_scores[clr_scores$Condition == g, ]

  cov_data <- cov.wt(
    cbind(group_data$NMDS1, group_data$NMDS2),
    wt = rep(1 / nrow(group_data), nrow(group_data))
  )$cov

  center <- c(mean(group_data$NMDS1), mean(group_data$NMDS2))
  ellipse_coords <- veganCovEllipse(cov_data, center)

  data.frame(NMDS1 = ellipse_coords[,1],
             NMDS2 = ellipse_coords[,2],
             Condition = g)
}))

clr_nMDS <- plot_ordination(
  ps_gut, GP.ord.clr,
  type = "samples",
  color = "Condition"
) +
  geom_path(data = df_ellipse_clr,
            aes(x = NMDS1, y = NMDS2, colour = Condition, group = Condition),
            size = 0.8) +
 scale_color_manual(values = c("Elevated CO2" = "#D4476C",
                              "Reference"    = "#2C728E")) +
  labs(color = "Condition") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(fill = NA, color = "black", linewidth = 1),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +
  annotate("text", x = Inf, y = Inf, label = stress_label,
           hjust = 1.05, vjust = 2.8, size = 4) +
  annotate("text", x = Inf, y = Inf, label = adonis_label,
           hjust = 1.05, vjust = 1.25, size = 4, parse = TRUE)

clr_nMDS

# Save
ggsave("Figures_Alpha_Beta_div/CLR_NMDS.png", clr_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
ggsave("Figures_Alpha_Beta_div/CLR_NMDS.svg", clr_nMDS, width = 15, height = 10, units = "cm", dpi = 900)

```

# Include water 
```{r}
library(compositions)
library(vegan)
library(phyloseq)
library(ggplot2)
library(dplyr)

# -------------------------
# Combine gut + water
# -------------------------

otu_all <- merge_phyloseq(
  otu_table(ps_gut),
  otu_table(ps_water)
)

# Taxonomy
tax_all <- tax_table(ps)[taxa_names(otu_all), ]

# -------------------------
# Metadata
# -------------------------

meta_gut <- data.frame(sample_data(ps_gut))
meta_water <- data.frame(sample_data(ps_water))

meta_all <- rbind(meta_gut, meta_water)

# Reorder metadata to exactly match OTU table
meta_all <- meta_all[match(sample_names(otu_all), rownames(meta_all)), , drop = FALSE]

# Check
all(rownames(meta_all) == sample_names(otu_all))

# Make phyloseq object WITHOUT a tree
ps_shikine <- phyloseq(
  otu_all,
  tax_all,
  sample_data(meta_all)
)

# Remove ASVs occurring in fewer than 2 samples
ps_shikine <- filter_taxa(
  ps_shikine,
  function(x) sum(x > 0) >= 2,
  prune = TRUE
)

# -------------------------
# CLR transform
# -------------------------

clr_transform <- function(x){
  x <- x + 1
  clr(x)
}

ps_clr_all <- transform_sample_counts(
  ps_shikine,
  clr_transform
)

# -------------------------
# NMDS (gut + water)
# -------------------------

GP.ord.clr_all <- ordinate(
  ps_clr_all,
  method = "NMDS",
  distance = "euclidean"
)

stress_val_all <- round(GP.ord.clr_all$stress, 3)
stress_label_all <- paste0("Stress = ", stress_val_all)

# -------------------------
# Euclidean distance
# -------------------------

clr_dist_all <- phyloseq::distance(
  ps_clr_all,
  method = "euclidean"
)

# -------------------------
# Metadata
# -------------------------

meta_all <- data.frame(
  sample_data(ps_clr_all)
)

# Make sure SampleType is a factor
meta_all$SampleType <- factor(
  meta_all$SampleType,
  levels = c("Sample", "H2O"),
  labels = c("Tissue", "Water")
)

# Condition
meta_all$Condition <- factor(
  meta_all$Condition,
  levels = c("Reference", "Elevated CO2")
)

#-------------------------
# PERMANOVA
#-------------------------

adonis_clr_all <- adonis2(
  clr_dist_all ~ SampleType + Condition,
  data = meta_all,
  permutations = 999,
  by = "margin"
)

adonis_clr_all

#            Df SumOfSqs      R2      F Pr(>F)    
#SampleType  1    35140 0.18968 9.1814  0.001 ***
#Condition   1    23998 0.12954 6.2702  0.001 ***

#-------------------------
# NMDS scores
#-------------------------

clr_scores_all <- as.data.frame(
  scores(GP.ord.clr_all, display = "sites")
)

clr_scores_all$Condition <- meta_all$Condition
clr_scores_all$SampleType <- meta_all$SampleType

# Group = Condition × SampleType
clr_scores_all$Group <- interaction(
  clr_scores_all$Condition,
  clr_scores_all$SampleType,
  sep = "_"
)

#-------------------------
# Ellipses
#-------------------------

df_ellipse_clr_all <- do.call(
  rbind,
  lapply(levels(clr_scores_all$Group), function(g){

    group_data <- clr_scores_all[
      clr_scores_all$Group == g,
    ]

    cov_data <- cov.wt(
      cbind(group_data$NMDS1,
            group_data$NMDS2),
      wt = rep(1 / nrow(group_data),
               nrow(group_data))
    )$cov

    center <- c(
      mean(group_data$NMDS1),
      mean(group_data$NMDS2)
    )

    ellipse_coords <- veganCovEllipse(
      cov_data,
      center
    )

    data.frame(
      NMDS1 = ellipse_coords[,1],
      NMDS2 = ellipse_coords[,2],
      Group = g,
      Condition = unique(group_data$Condition),
      SampleType = unique(group_data$SampleType)
    )
  })
)

#-------------------------
# Plot
#-------------------------

clr_nMDS_all <- ggplot(
  clr_scores_all,
  aes(
    NMDS1,
    NMDS2,
    colour = Condition,
    shape = SampleType
  )
) +

  geom_path(
    data = df_ellipse_clr_all,
    aes(
      x = NMDS1,
      y = NMDS2,
      group = Group,
      colour = Condition
    ),
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +

  geom_point(size = 1.5) +

  scale_color_manual(
    values = c(
      "Reference" = "#2C728E",
      "Elevated CO2" = "#D4476C"
    ),
    labels = c(
      "Reference",
      expression("Elevated CO"[2])
    )
  ) +

  scale_shape_manual(
    values = c(
      Tissue = 16,
      Water = 17
    )
  ) +

  labs(
    colour = "Site",
    shape = "Sample Type"
  ) +

  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(
      fill = NA,
      color = "black",
      linewidth = 1
    ),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +

  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = stress_label_all,
    hjust = 1.05,
    vjust = 4,
    size = 4.1
  ) +

  annotate(
  "text",
  x = Inf,
  y = Inf,
  label = "'Type:'~~italic(F)==9.18*','~~italic(p)<0.001~'***'",
  hjust = 1.05,
  vjust = 1.25,
  size = 4,
  parse = TRUE
) +

annotate(
  "text",
  x = Inf,
  y = Inf,
  label = "'Site:'~~italic(F)==6.27*','~~italic(p)<0.001~'***'",
  hjust = 1.05,
  vjust = 2.3,
  size = 4,
  parse = TRUE
)

clr_nMDS_all

ggsave(
  filename = "Figures_Alpha_Beta_div/Tissue_and_Water_NMDS.svg",
  plot = clr_nMDS_all,
  width = 7,
  height = 4,
  device = "svg"
)

#Amplicon sequence variants present in fewer than two samples were removed prior to CLR #transformation and Aitchison distance analyses to reduce the influence of extremely sparse 
#taxa.

head(sample_names(otu_all))
head(rownames(meta_all))

setdiff(sample_names(otu_all), rownames(meta_all))
setdiff(rownames(meta_all), sample_names(otu_all))
```

# philr -- Phylogenetic Isometric Log-Ratio Transform. This package provides functions for the analysis of compositional data (e.g., data representing proportions of different variables/parts). Specifically this package allows analysis of compositional data where the parts can be related through a phylogenetic tree (as is common in microbiota survey data) and makes available the Isometric Log Ratio transform built from the phylogenetic tree and utilizing a weighted reference measure.

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("philr")
```

```{r}
library(philr)
library(ape)
library(vegan)
library(phyloseq)

# Keep taxa with >10 total counts and present in at least 3 samples
ps_filt <- filter_taxa(
  ps_gut,
  function(x) sum(x) > 10 & sum(x > 0) >= 3,
  prune = TRUE
)

cat("Remaining taxa:", ntaxa(ps_filt), "\n")
cat("Remaining samples:", nsamples(ps_filt), "\n")

# -------------------------------------------------
# 1. FIX TREE ↔ ASV NAME MISMATCH (robust + automatic)
# -------------------------------------------------

tree <- phy_tree(ps_filt)
taxa_ps <- taxa_names(ps_filt)

# If tree tips are DNA sequences instead of ASV IDs,
# automatically rename them to ASV names using order matching
if(!any(tree$tip.label %in% taxa_ps)){
  
  message("Renaming tree tips to ASV IDs...")
  
  # get OTU table taxa order
  otu_taxa <- taxa_names(ps_filt)
  
  # make mapping: sequence order -> ASV names
  # assumes tree was built from the same sequence set
  if(length(tree$tip.label) != length(otu_taxa)){
    stop("Tree and OTU table have different numbers of taxa!")
  }
  
  # rename tree tips to match ASV IDs
  tree$tip.label <- otu_taxa
}

# prune tree to exact taxa present
tree_pruned <- keep.tip(tree, taxa_ps)
phy_tree(ps_filt) <- tree_pruned

# final safety check
if(!setequal(taxa_names(ps_filt), phy_tree(ps_filt)$tip.label)){
  stop("Tree tips and taxa names still do not match!")
}

message("Tree and taxa successfully matched ✔")

# -------------------------------------------------
# 2. Run PhILR transform
# -------------------------------------------------
# -------------------------------------------------
# Prepare inputs for philr (SAFE METHOD)
# -------------------------------------------------

# Extract count matrix
otu <- as(otu_table(ps_filt), "matrix")

# philr requires samples as rows, taxa as columns
if(taxa_are_rows(ps_filt)){
  otu <- t(otu)
}

# Extract tree
tree <- phy_tree(ps_filt)

# Final safety check
stopifnot(all(colnames(otu) == tree$tip.label))

# -------------------------------------------------
# Run PhILR
# -------------------------------------------------

philr_ps <- philr(
  otu,
  tree,
  part.weights = "enorm.x.gm.counts",
  ilr.weights  = "blw.sqrt",
  return.all   = FALSE,
  pseudocount  = 0.5
)

# Euclidean distance
philr_dist <- dist(philr_ps, method = "euclidean")

# NMDS
GP.ord.philr <- metaMDS(philr_dist, trymax = 100)

# Extract NMDS scores
philr_scores <- as.data.frame(scores(GP.ord.philr, display = "sites"))

# Add sample IDs explicitly
philr_scores$SampleID <- rownames(philr_scores)

# Match metadata safely
meta <- data.frame(sample_data(ps_filt))
meta$SampleID <- rownames(meta)

philr_scores <- merge(philr_scores, meta, by = "SampleID")

philr_scores$Condition <- factor(philr_scores$Condition)

# -------------------------------------------------
# 3. PERMANOVA
# -------------------------------------------------

adonis_philr <- adonis2(
  philr_dist ~ Condition,
  data = data.frame(sample_data(ps_filt)),
  permutations = 999
)

F_val <- round(adonis_philr$F[1], 2)
p_val <- adonis_philr$`Pr(>F)`[1]


adonis_label <- expression(italic(F) == 16.77 * "," ~ italic(p) < 0.001 ~ "***")

stress_val <- round(GP.ord.philr$stress, 3)
stress_label <- paste0("Stress = ", stress_val)

# -------------------------------------------------
# 4. Ellipses
# -------------------------------------------------

df_ellipse_philr <- do.call(rbind, lapply(levels(philr_scores$Condition), function(g){
  group_data <- philr_scores[philr_scores$Condition == g, ]
  cov_data <- cov.wt(cbind(group_data$NMDS1, group_data$NMDS2),
                     wt = rep(1 / nrow(group_data), nrow(group_data)))$cov
  center <- c(mean(group_data$NMDS1), mean(group_data$NMDS2))
  ellipse_coords <- veganCovEllipse(cov_data, center)
  data.frame(NMDS1 = ellipse_coords[,1],
             NMDS2 = ellipse_coords[,2],
             Condition = g)
}))

# -------------------------------------------------
# 5. Plot
# -------------------------------------------------

philr_nMDS <- ggplot(philr_scores, aes(NMDS1, NMDS2, colour = Condition)) +
  geom_point(size = 1.5) +
  geom_path(data = df_ellipse_philr,
            aes(x = NMDS1, y = NMDS2, colour = Condition, group = Condition),
            size = 0.8) +
 scale_color_manual(values = c("Elevated CO2" = "#D4476C",
                              "Reference"    = "#2C728E")) +
  labs(color = "Condition") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(fill = NA, color = "black", linewidth = 1),
    axis.line = element_blank(),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  ) +
  annotate("text", x = Inf, y = Inf, label = stress_label,
           hjust = 1.05, vjust = 2.8, size = 4) +
  annotate("text", x = Inf, y = Inf, label = adonis_label,
           hjust = 1.05, vjust = 1.25, size = 4, parse = TRUE)

philr_nMDS

# Save
ggsave("Figures_Alpha_Beta_div/PhilR_NMDS.png", philr_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
ggsave("Figures_Alpha_Beta_div/PhilR_NMDS.svg", philr_nMDS, width = 15, height = 10, units = "cm", dpi = 900)
```

# All beta plots
```{r}
all.beta.plots <- nMDS + w_unifrac_nMDS + unifrac_nMDS + clr_nMDS + philr_nMDS + plot_layout(nrow = 1)

# This combo says: this pattern is robust across abundance, phylogeny, and compositional methods.
main.beta.plots <- nMDS + w_unifrac_nMDS + philr_nMDS + plot_layout(nrow = 1)

# Supporting metrics: Unweighted UniFrac (Presence/absence version of phylogenetic signal)
# CLR Aitchison (Compositional without phylogeny)
supp.beta.plots <- unifrac_nMDS + clr_nMDS + plot_layout(nrow = 1)
```

# Patchwork plotting Alpha + Beta Plots
```{r}
library(patchwork)

# -----------------------------
# Remove legends from alpha row
# -----------------------------
combined_alpha_plots_nolegend <- combined_alpha_plots +
  theme(legend.position = "none")

# =====================================================
# MAIN FIGURE
# CLR + Weighted UniFrac + PhILR
# =====================================================

clr_main_noleg        <- clr_nMDS + theme(legend.position = "none")
w_unifrac_main_noleg  <- w_unifrac_nMDS + ylab(NULL) + theme(legend.position = "none")
philr_main_legend     <- philr_nMDS + ylab(NULL)   # keep legend here

main_beta_row <- clr_main_noleg +
                 w_unifrac_main_noleg +
                 philr_main_legend +
  plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")

# Combine alpha + beta (MAIN FIG)
main_alpha_beta_fig <- combined_alpha_plots_nolegend / main_beta_row +
  plot_layout(nrow = 2, heights = c(1, 1))

main_alpha_beta_fig

# Save MAIN
ggsave("Figures_Alpha_Beta_div/Fig_Main_Alpha_Beta.png", plot = main_alpha_beta_fig,
       width = 24, height = 24, units = "cm", dpi = 900)

ggsave("Figures_Alpha_Beta_div/Fig_Main_Alpha_Beta.svg", plot = main_alpha_beta_fig,
       width = 24, height = 24, units = "cm", dpi = 900)


# =====================================================
# SUPPLEMENTARY FIGURE
# Bray–Curtis + Unweighted UniFrac
# =====================================================

bray_supp_noleg     <- Bray_nMDS + theme(legend.position = "none")
unifrac_supp_legend <- UW_unifrac_nMDS + ylab(NULL)

supp_beta_fig <- bray_supp_noleg +
                 unifrac_supp_legend +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")

supp_beta_fig

# Save SUPPLEMENT
ggsave("Figures_Alpha_Beta_div/Fig_Supp_Beta_Robustness.png", plot = supp_beta_fig,
       width = 18, height = 10, units = "cm", dpi = 900)

ggsave("Figures_Alpha_Beta_div/Fig_Supp_Beta_Robustness.svg", plot = supp_beta_fig,
       width = 18, height = 10, units = "cm", dpi = 900)
```

```{r}
# =========================================================
# Shared microbial taxa: Gut vs Water and Reference vs Elevated CO2
# =========================================================

library(phyloseq)
library(dplyr)
#install.packages("eulerr")
library(eulerr)

# ---------------------------------------------------------
# Keep only gut + water samples
# ---------------------------------------------------------

ps_venn <- subset_samples(
  ps_shikine,
  SampleType %in% c("Sample", "H2O")
)

ps_venn <- prune_taxa(
  taxa_sums(ps_venn) > 0,
  ps_venn
)

# ---------------------------------------------------------
# Check sample numbers
# ---------------------------------------------------------

table(
  sample_data(ps_venn)$SampleType,
  sample_data(ps_venn)$Condition
)

# ---------------------------------------------------------
# Split into four groups
# ---------------------------------------------------------

ps_ref_gut <- subset_samples(
  ps_venn,
  SampleType == "Sample" &
    Condition == "Reference"
)

ps_ref_water <- subset_samples(
  ps_venn,
  SampleType == "H2O" &
    Condition == "Reference"
)

sample_data(ps_ref_water)

ps_high_gut <- subset_samples(
  ps_venn,
  SampleType == "Sample" &
    Condition == "Elevated CO2"
)

ps_high_water <- subset_samples(
  ps_venn,
  SampleType == "H2O" &
    Condition == "Elevated CO2"
)

sample_data(ps_high_water)

# Remove taxa absent from each subset
ps_ref_gut <- prune_taxa(taxa_sums(ps_ref_gut) > 0, ps_ref_gut)
ps_ref_water <- prune_taxa(taxa_sums(ps_ref_water) > 0, ps_ref_water)
ps_high_gut <- prune_taxa(taxa_sums(ps_high_gut) > 0, ps_high_gut)
ps_high_water <- prune_taxa(taxa_sums(ps_high_water) > 0, ps_high_water)

# ---------------------------------------------------------
# Sample numbers
# ---------------------------------------------------------

nsamples(ps_ref_gut)
nsamples(ps_ref_water)
nsamples(ps_high_gut)
nsamples(ps_high_water)

# ---------------------------------------------------------
# Identify core ASVs
# Core = present in 100% of samples in that group
# ---------------------------------------------------------

get_core_asvs <- function(ps_obj) {

  otu <- as.data.frame(otu_table(ps_obj))

  if (!taxa_are_rows(ps_obj)) {
    otu <- t(otu)
  }

  prevalence <- rowSums(otu > 0)

  names(
    prevalence[
      prevalence == nsamples(ps_obj)
    ]
  )
}

core_ref_gut <- get_core_asvs(ps_ref_gut)
core_ref_water <- get_core_asvs(ps_ref_water)
core_high_gut <- get_core_asvs(ps_high_gut)
core_high_water <- get_core_asvs(ps_high_water)

# Check numbers
sapply(
  list(
    Reference_Gut = core_ref_gut,
    Reference_Water = core_ref_water,
    Elevated_Gut = core_high_gut,
    Elevated_Water = core_high_water
  ),
  length
)

# =========================================================
# Euler diagrams
# =========================================================

library(eulerr)

# ---------------------------------------------------------
# Colours
# ---------------------------------------------------------

gut_fill <- c(
  Reference = "#2C728E",
  Elevated = "#D4476C"
)

gut_line <- c(
  Reference = "#1F4E64",
  Elevated = "#9B3654"
)

water_fill <- c(
  Reference = "#0080BFFF",
  Elevated = "#CC0000FF"
)

water_line <- c(
  Reference = "#004C73FF",
  Elevated = "#7A0000FF"
)


# =========================================================
# 1. Reference: Gut vs Water
# =========================================================

ref_fit <- euler(
  list(
    Gut = core_ref_gut,
    Water = core_ref_water
  )
)

plot(
  ref_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      water_fill["Reference"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      water_line["Reference"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Reference: Gut vs Water"
)

svg(
  "Figures_Alpha_Beta_div/Euler_Reference_Gut_vs_Water.svg",
  width = 6,
  height = 6
)

plot(
  ref_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      water_fill["Reference"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      water_line["Reference"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Reference: Gut vs Water"
)

dev.off()


# =========================================================
# 2. Elevated CO2: Gut vs Water
# =========================================================

high_fit <- euler(
  list(
    Gut = core_high_gut,
    Water = core_high_water
  )
)

plot(
  high_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Elevated"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Elevated"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Elevated CO2: Gut vs Water"
)

svg(
  "Figures_Alpha_Beta_div/Euler_Elevated_Gut_vs_Water.svg",
  width = 6,
  height = 6
)

plot(
  high_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Elevated"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Elevated"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Elevated CO2: Gut vs Water"
)

dev.off()


# =========================================================
# 3. Gut: Reference vs Elevated CO2
# =========================================================

gut_fit <- euler(
  list(
    Reference = core_ref_gut,
    Elevated = core_high_gut
  )
)

plot(
  gut_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      gut_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      gut_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Gut: Reference vs Elevated CO2"
)

svg(
  "Figures_Alpha_Beta_div/Euler_Gut_Reference_vs_Elevated.svg",
  width = 6,
  height = 6
)

plot(
  gut_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      gut_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      gut_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Gut: Reference vs Elevated CO2"
)

dev.off()


# =========================================================
# 4. Water: Reference vs Elevated CO2
# =========================================================

water_fit <- euler(
  list(
    Reference = core_ref_water,
    Elevated = core_high_water
  )
)

plot(
  water_fit,
  fills = list(
    fill = unname(c(
      water_fill["Reference"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      water_line["Reference"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Water: Reference vs Elevated CO2"
)

svg(
  "Figures_Alpha_Beta_div/Euler_Water_Reference_vs_Elevated.svg",
  width = 6,
  height = 6
)

plot(
  water_fit,
  fills = list(
    fill = unname(c(
      water_fill["Reference"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      water_line["Reference"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Water: Reference vs Elevated CO2"
)

dev.off()

venn_panel <- (venn_ref_plot | venn_high_plot) /
              (venn_gut_sites | venn_water_sites)

ggsave("Figures_Alpha_Beta_div/combined_venn_diagrams.svg", venn_panel,
       width = 18, height = 18, units = "cm", dpi = 900)
```

```{r}
# =========================================================
# Euler diagrams
# =========================================================

library(eulerr)
library(gridExtra)

# ---------------------------------------------------------
# Colours
# ---------------------------------------------------------

gut_fill <- c(
  Reference = "#2C728E",
  Elevated = "#D4476C"
)

gut_line <- c(
  Reference = "#1F4E64",
  Elevated = "#9B3654"
)

water_fill <- c(
  Reference = "#0080BFFF",
  Elevated = "#CC0000FF"
)

water_line <- c(
  Reference = "#004C73FF",
  Elevated = "#7A0000FF"
)


# =========================================================
# Make Euler fits
# =========================================================

ref_fit <- euler(
  list(
    Gut = core_ref_gut,
    Water = core_ref_water
  )
)

high_fit <- euler(
  list(
    Gut = core_high_gut,
    Water = core_high_water
  )
)

gut_fit <- euler(
  list(
    Reference = core_ref_gut,
    Elevated = core_high_gut
  )
)

water_fit <- euler(
  list(
    Reference = core_ref_water,
    Elevated = core_high_water
  )
)


# =========================================================
# 1. Reference: Gut vs Water
# =========================================================

svg(
  "Figures_Alpha_Beta_div/Euler_Reference_Gut_vs_Water.svg",
  width = 6,
  height = 6
)

plot(
  ref_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      water_fill["Reference"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      water_line["Reference"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Reference: Gut vs Water"
)

dev.off()


# =========================================================
# 2. Elevated CO2: Gut vs Water
# =========================================================

svg(
  "Figures_Alpha_Beta_div/Euler_Elevated_Gut_vs_Water.svg",
  width = 6,
  height = 6
)

plot(
  high_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Elevated"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Elevated"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Elevated CO2: Gut vs Water"
)

dev.off()


# =========================================================
# 3. Gut: Reference vs Elevated CO2
# =========================================================

svg(
  "Figures_Alpha_Beta_div/Euler_Gut_Reference_vs_Elevated.svg",
  width = 6,
  height = 6
)

plot(
  gut_fit,
  fills = list(
    fill = unname(c(
      gut_fill["Reference"],
      gut_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      gut_line["Reference"],
      gut_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Gut: Reference vs Elevated CO2"
)

dev.off()


# =========================================================
# 4. Water: Reference vs Elevated CO2
# =========================================================

svg(
  "Figures_Alpha_Beta_div/Euler_Water_Reference_vs_Elevated.svg",
  width = 6,
  height = 6
)

plot(
  water_fit,
  fills = list(
    fill = unname(c(
      water_fill["Reference"],
      water_fill["Elevated"]
    )),
    alpha = 0.4
  ),
  edges = list(
    col = unname(c(
      water_line["Reference"],
      water_line["Elevated"]
    )),
    lwd = 1
  ),
  labels = TRUE,
  quantities = TRUE,
  main = "Water: Reference vs Elevated CO2"
)

dev.off()


# =========================================================
# Combined 4-panel figure
# =========================================================

svg(
  "Figures_Alpha_Beta_div/combined_venn_diagrams.svg",
  width = 12,
  height = 12
)

grid.arrange(

  plot(
    ref_fit,
    fills = list(
      fill = unname(c(
        gut_fill["Reference"],
        water_fill["Reference"]
      )),
      alpha = 0.4
    ),
    edges = list(
      col = unname(c(
        gut_line["Reference"],
        water_line["Reference"]
      )),
      lwd = 1
    ),
    labels = TRUE,
    quantities = TRUE,
    main = "Reference: Gut vs Water"
  ),

  plot(
    high_fit,
    fills = list(
      fill = unname(c(
        gut_fill["Elevated"],
        water_fill["Elevated"]
      )),
      alpha = 0.4
    ),
    edges = list(
      col = unname(c(
        gut_line["Elevated"],
        water_line["Elevated"]
      )),
      lwd = 1
    ),
    labels = TRUE,
    quantities = TRUE,
    main = "Elevated CO2: Gut vs Water"
  ),

  plot(
    gut_fit,
    fills = list(
      fill = unname(c(
        gut_fill["Reference"],
        gut_fill["Elevated"]
      )),
      alpha = 0.4
    ),
    edges = list(
      col = unname(c(
        gut_line["Reference"],
        gut_line["Elevated"]
      )),
      lwd = 1
    ),
    labels = TRUE,
    quantities = TRUE,
    main = "Gut: Reference vs Elevated CO2"
  ),

  plot(
    water_fit,
    fills = list(
      fill = unname(c(
        water_fill["Reference"],
        water_fill["Elevated"]
      )),
      alpha = 0.4
    ),
    edges = list(
      col = unname(c(
        water_line["Reference"],
        water_line["Elevated"]
      )),
      lwd = 1
    ),
    labels = TRUE,
    quantities = TRUE,
    main = "Water: Reference vs Elevated CO2"
  ),

  ncol = 2
)

dev.off()
```







